---
title: "DAY2 - plotting"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This document contains the code that will be displayed during the second day of the course. It also includes extra code for better understanding of the concepts.


### Dataset: superhero
We will use the hero dataset throughout the document.

There are two files related to superhero. first dataset contains various information about the superhero’s appearance,their body measurements etc. There are 10 variables in this dataset. Variable descriptions are as follows:
Name: Name of the superhero
Gender: Gender of the superhero
Eye color: Eye color of the superhero
Race: Species of the superhero
Height: Height of the superhero (in cms)
Publisher: Comic category of the superhero
Skin color: Skin color of the superhero
Alignment: Superhero’s nature.
Weight: Weight of the superhero (in kgs)

The second dataset will be used on the third day of the course and records the superhero attributes such as their strength level,speed level etc.

## Import data and make a few changes to better visualise the data later
```{r}
library(tidyverse)
heroes = read.csv("../datasets/heroes_information.csv")
View(heroes)
str(heroes)
heroes$Publisher = as.factor(heroes$Publisher) # change Publisher variable to factor
heroes$Gender = as.factor(heroes$Gender) # change Gender variable to factor
levels(heroes$Gender)= c("Not Specified", "Female", "Male")# Define a new level as "Not Specified"


```


# Missing values
```{r}
# Let´s create a vector with multiple NAs
my_data = c(3, 4, NA, 4, 5, NA, 1, NA, NA)

# Test if there are NAs in our data
is.na(my_data) # Gives us a logical vector
which(is.na(my_data)) # gives us the location (indexes) of these NAs (TRUE)

# However if we are testing a large vector
sum(is.na(my_data)) # identify number of NA
table(is.na(my_data)) # counts number of NA

# We can test a column of our hero dataset
which(is.na(heroes$Weight))

# Remove NA
remove_na = is.na(my_data) # logical vector which information about where NAs are
my_data = my_data[!remove_na] # subset my_data to not include the TRUE elements
my_data
is.na(my_data)

# Now let's work with a data frame
data.frame = data.frame (c1 = 1:8, c2 = factor (c("a", "b", "a", "c", "b", "c", "a", "b")))
View(data.frame)
data.frame[4,1] = data.frame[6,2] = NA

# Test if there are NAs
is.na(data.frame)
table(is.na(data.frame))

# Filter missing data
na.omit(data.frame)

View(data.frame)
```

# Recoding missing values
In our dataset heroes we can see that a lot of missing values are actually treated as -. We can recode this by using normal subsettings and assignment operations. 

```{r}
# Let's work on a similar data.frame
data.frame = data.frame (c1 = 1:8, c2 = factor (c("a", "-", "a", "c", "b", "c", "-", "b")))
View(data.frame)

# Check if there is any NA
is.na(data.frame)

# However we know that column 2 misses some values as - 
sum(data.frame == "-") 

# Let´s recode them to NAs
data.frame[data.frame == "-"] = NA
is.na(data.frame)
View(data.frame)

# The hero dataset has some missing values that are not strictly shown as NA. Can you identify any and recode them?


```


## Merging data
When you have related information in two or more data.frames, you can use the `merge()` function to join them correctly. The two data.frames that you want to join must have one or more columns with the same name - these are the ID columns that R uses to reorder the rows to match in both data.frames. If there are equivalent columns but the names are different, the column name in one has to be renamed to match the name in the other data.frame (as in the following example).

```{r}
# heroes <- read.csv("../datasets/heroes_information.csv")
powers <- read.csv("../datasets/super_hero_powers.csv")

colnames(powers) <- sub("hero_names", "name", colnames(powers))

# see the difference between the following
heroes_with_powers <- merge(heroes, powers, by = "name")
heroes_with_powers <- merge(heroes, powers, by = "name", all = TRUE)
heroes_with_powers <- merge(heroes, powers, by = "name", all.x = TRUE)
heroes_with_powers <- merge(heroes, powers, by = "name", all.y = TRUE)

```

Note that there are duplicated names in the `heroes` dataset. Because of this, the merging may give unexpected results. Try doing the same as the above, after removing the duplicated superheroes. Hint: you can filter out the duplicated superheroes using the `duplicated()` function on the `name` column.

```{r}
# Merge after removing duplicates

```


## Reshaping data
Sometimes the data is not in the right format for the plot or analysis you want to run. You may need to reshape the data into wide or long format. The `tidyr` package (part of tidyverse) has the `pivot_` functions for this. An alternative is the `data.table` package with the `melt` and `dcast` functions.

```{r}
powers %>%
  pivot_longer(cols = -name, names_to = "power", values_to = "value") -> powers_long

powers_long %>%
  pivot_wider(names_from = "power", values_from = "value") -> powers_wide

```

## Basic statistics

Are the mean weights of male and female superheroes different? Is this difference statistically significant? Let's set our significance cutoff at 0.5

```{r}
male_heroes_weight <- heroes[heroes$Gender == "Male", "Weight"]
female_heroes_weight <- heroes[heroes$Gender == "Female", "Weight"]

t.test(male_heroes_weight, female_heroes_weight)
```

Is the weight of a superhero correlated to their height?
```{r}
cor.test(heroes$Weight, heroes$Height)

```



## DAY 2 AFTERNOON SESSIONS: PLOTTING WITH BASE R AND GGPLOT2

## Base R
Let´s start by exploring basic R plotting. 
It is important to understand there is no need to remember all the arguments and optional things we can do with graphs. The important is to understand how plotting works and know the sources to get the information to reach your desired plot.

```{r}
# Histograms
hist(heroes$Height) # Histogram to see distribution
hist(heroes$Height, breaks=50) # More detail
hist(heroes$Height, breaks=50, col="red") # Add colour
hist(heroes$Height, breaks=50, col="red", main = "Height distribution") # Add title

# Barplots
plot(heroes$Publisher) # Barplot
plot(heroes$Publisher, las=2) # turn 90 degrees x labels
plot(heroes$Publisher, las=2, cex.names=0.6) # change size labels
plot(heroes$Publisher, las=2, cex.names=0.6, ylim= c(0,500)) # Increase y axis range
plot(heroes$Publisher, las=2, cex.names=0.6, ylim= c(0,500), main = "Barplot of heroes by publisher") # add title


# XY graphs / scatterplots / correlation
plot(heroes$Height, heroes$Weight) # Scatterplot of x and y values
plot(heroes$Height, heroes$Weight, col="blue") #colour for every gender
plot(heroes$Height, heroes$Weight, col=heroes$Gender) #colour for every gender
plot(heroes$Height, heroes$Weight, col=heroes$Gender, pch=24) #change symbol
plot(heroes$Height, heroes$Weight, col=heroes$Gender, pch=24, cex=2) #change size

plot(heroes$Height, heroes$Weight, col=heroes$Gender, pch=24)
legend(x=700, y=700, legend = levels(heroes$Gender), col=c(1:3), pch= 24) #create legend

# Remember we can check the different symbols we can use  
ggpubr::show_point_shapes()

# Boxplots
boxplot(heroes$Height) # We can see distribution of the variable Height
boxplot(heroes$Height,heroes$Weight) # We can plot multiple boxplots (multiple variables)
boxplot(heroes$Height,heroes$Weight, names=c("Height", "Weight")) # Add name of variables plotted

# If we want to plot data based on a categorical data we can use plot()
plot(heroes$Gender, heroes$Height) # plot height data splited by Gender
plot(heroes$Gender, heroes$Height, col=c("green","blue", "red")) #add colour
plot(heroes$Gender, heroes$Height, col=c("green","blue", "red"), xlab = "Gender" ) #change x axis title

# Save graphs
pdf("Barplot_heroes.pdf")
plot(heroes$Publisher, las=2, cex.names = 0.6, ylim= c(0,500), main = "Barplot of heroes by publisher") 
dev.off()

pdf("Barplot_heroes.pdf", width=15, height = 15)
plot(heroes$Publisher, las=2, cex.names = 0.6, ylim= c(0,500), main = "Barplot of heroes by publisher") 
dev.off()

png("Hero_scatterplot.png")
plot(heroes$Height, heroes$Weight, col=heroes$Gender, pch=24)
dev.off()

# More advanced information
# Colours can be defined in different ways

# specifing colour name (most common)
boxplot(heroes$Height, col="blue") 
# you can find all the names for colours that R offers using
colors()

# specifing hexadecimal color code 
boxplot(heroes$Height, col="#FF00CC") 
# https://htmlcolorcodes.com will allow you to pick a colour and get the code

# using number 
boxplot(heroes$Height, col=431) 

# using rgb() function to build a color (red, green, blue)
boxplot(heroes$Height, col=rgb(0.3, 0.7, 0.3))
```
# Exercises

```{r}
# Plot a boxplot of the Weight by Alignment.


```



```{r}
# Plot the frequency of every alignment in the dataset


```


```{r}
# Colour the previous plot



```



## ggplot2
Plotting package that is part of tidyverse.
We will work mostly on scatterplot

```{r}
# Install/load ggplot2
#install.packages("ggplot2")
library(ggplot2)


# Let´s get rid of the not specified group for the sake of better visualisation
heroes = heroes[heroes$Gender != "Not Specified",]


# Plot the canvas
ggplot() 

# Add the data and mapping information
ggplot(data = heroes, mapping = aes(Weight,Height))

# Add the geometry
ggplot(data = heroes, mapping = aes(Weight,Height))+
	geom_point()

# Data and mapping can be given as global or on the layer. Remember that will affect the inheritance and where things apply.
ggplot(data = heroes)+
	geom_point(mapping = aes(Weight,Height))

ggplot()+
	geom_point(mapping = aes(Weight,Height), data= heroes)


# Use different type of gemoetry
ggplot(data = heroes, mapping = aes(Weight,Height)) +
	geom_line() 

# We can add different geometries too
ggplot(data = heroes, mapping = aes(Weight,Height)) +
  geom_point() +
	geom_line() 

# There is really no need to specify that the arguments are the data or mapping.
ggplot(heroes, aes(Weight,Height))+
	geom_point()

# We can control the style with themes that are defined already
ggplot(data = heroes, mapping = aes(Weight,Height)) +
	geom_point(aes(colour = Gender)) +
	theme_classic() # classic style


ggplot(data = heroes, mapping = aes(Weight,Height))+
	geom_point(aes(colour = Gender))+
	theme_dark() # dark style


# Split data into different graphs using facets
ggplot(data = heroes, mapping = aes(Weight,Height)) +
	geom_point(aes(colour = Gender)) +
	theme_dark() +
	facet_wrap(~Alignment)


ggplot(data = heroes, mapping = aes(Weight,Height)) +
	geom_point(aes(colour = Gender)) +
	theme_dark() + 
	facet_grid(~Alignment)

ggplot(data = heroes, mapping = aes(Weight,Height))+
	geom_point(aes(colour = Gender))+
	theme_dark()+
	facet_grid(Gender~Alignment) # it can be split in two variable


```

Let´s work now a bit on a different type of graph: boxplot

```{r}
# Boxplot for Gender and Weight
ggplot(data = heroes, mapping = aes(Gender,Weight)) +
  geom_boxplot(aes(colour=Gender)) +
  theme_classic()

ggplot(data = heroes, mapping = aes(Gender,Weight)) +
  geom_boxplot(aes(colour=Gender))+
  theme_classic()+
  facet_grid(~Alignment)


```
Let´s learn how to save the plot as a variable and how to export it:

```{r}
# Save as a an object that we can print
myplot = ggplot(data = heroes, mapping = aes(Gender, Weight)) +
  geom_boxplot(aes(colour=Gender)) +
  theme_classic() +
  facet_grid(~Alignment) +
  ggtitle ("myplot")

print(myplot)

# Export the plot
ggsave("myplot.pdf", 
       plot = myplot, width = 10, height = 5)

ggsave("myplot.png", 
       plot = myplot)

ggsave("myplot.svg", 
       plot = myplot)

```



# Some things to have in mind  

```{r}
# Let´s take this one graph we have worked with before
ggplot(data = heroes, mapping = aes(Weight,Height)) +
	geom_point()

# Let's add some colour
ggplot(data = heroes, mapping = aes(Weight,Height)) +
	geom_point(colour = "Blue") 

# Different way
ggplot(data = heroes) + 
	geom_point( mapping = aes(Weight,Height),colour = "Blue") 

# Be careful! Not the same as
ggplot(data = heroes)+
	geom_point( mapping = aes(Weight,Height,colour = "Blue")) #If we add the colour inside aesthetic then it is linked to data and it is not the same!

# Colour based on a variable
ggplot(data = heroes, mapping = aes(Weight,Height))+
	geom_point(aes(colour = Gender)) # Here it makes sense as colour is linked now to data

# Specifiy colours
ggplot(data = heroes, mapping = aes(Weight,Height))+
	geom_point(aes(colour = Weight > 100)) # we can even decide a condition to based our colour in (we will see conditions later in the course)

```

Some geometries only need a single mapping

```{r}
ggplot(heroes) + 
  geom_histogram(aes(x = Weight))
```


#### Some practice

Modify the code below to make points as squares and more transparent.

```{r}
ggplot(heroes) + 
  geom_point(aes(Weight,  Height))


# Symbols as squares


# Symbols transparent



```

Hint 1: transparency is controlled with alpha, and shape with shape
Hint 2: remember the difference between mapping and setting aesthetics

```{r}
# Solution
ggplot(heroes) + 
  geom_point(aes(Weight,  Height), shape = "square", alpha=0.5)
```


Colour the data on the graph depending on alignment

```{r}
ggplot()
```


```{r}
# Solution
ggplot(heroes) + 
  geom_point(aes(Weight,  Height, colour= Alignment), shape = "square", alpha=0.5)
```

Now add a main title
```{r}
ggplot()
```


```{r}
# Solution
ggplot(heroes) + 
  geom_point(aes(Weight,  Height, colour= Alignment), shape = "square", alpha=0.5) +
  ggtitle ("Heroes: weight vs height by alignment")
```

Now let´s work with categorical data and boxplot.
First plot just the continuos data Weigth

```{r}
ggplot(data = heroes, mapping = aes(Weight)) +
  geom_boxplot()
```

We can see the boxplot representing the weights of the whole dataset. With a line there is the median represented.
Let´s add now the categorical data we would like to plot too.

```{r}
ggplot(data = heroes, mapping = aes(Weight,Gender)) +
  geom_boxplot()
```

Let´s consider just Male and Female data
```{r}

ggplot(data = heroes, mapping = aes(Weight,Gender)) +
  geom_boxplot()
```

Try to change the axis orientation so that Gender shows in x axis

```{r}
ggplot(data = heroes, mapping = aes(Gender, Weight)) +
  geom_boxplot() 
```


Let´s add the classic theme

```{r}
ggplot(data = heroes, mapping = aes(Gender, Weight)) +
  geom_boxplot() +
  theme_classic()
```

Add some colour
```{r}
ggplot(data = heroes, mapping = aes(Gender, Weight, colour= Gender)) +
  geom_boxplot()
```

Choose your own colours

```{r}
ggplot(data = heroes, mapping = aes(Gender, Weight, fill= Gender)) +
  geom_boxplot() +
  scale_fill_manual(values=c("red", "yellow"))


ggplot(data = heroes, mapping = aes(Gender, Weight)) +
  geom_boxplot(aes(fill= Gender)) +
  scale_fill_manual(values=c("green", "purple"))

```


Let us have a look to the following histogram

```{r}
ggplot(heroes, aes(Weight)) +
  geom_histogram()
```


Colour the bars in blue:

```{r}
ggplot(heroes, aes(Weight)) +
  geom_histogram(colour="Blue")
```

Try fill instead of colour to define the blue bars:

```{r}
ggplot(heroes, aes(Weight)) +
  geom_histogram(fill="Blue")
```

Going back to our x y plot:

```{r}
ggplot(heroes) + 
  geom_point(aes(Weight,  Height, colour= Alignment), shape = "square", alpha=0.5) +
  ggtitle ("Heroes: weight vs height by alignment")
```


Let´s add a theme:
```{r}
ggplot(heroes) + 
  geom_point(aes(Weight,  Height, colour= Alignment), shape = "square", alpha=0.5) +
  ggtitle ("Heroes: weight vs height by alignment") +
  theme_minimal()
```

Now we will split the data using facets.
First let´s try facet_wrap and split the data by alignment.

```{r}
ggplot(heroes) + 
  geom_point(aes(Weight,  Height, colour= Alignment), shape = "square", alpha=0.5) +
  ggtitle ("Heroes: weight vs height by alignment") +
  theme_minimal() +
  facet_wrap(~Alignment)
```

Let´s try facet_grid using Gender and Alignment to split the data.

```{r}
ggplot(heroes) + 
  geom_point(aes(Weight,  Height, colour= Alignment), shape = "square", alpha=0.5) +
  ggtitle ("Heroes: weight vs height by alignment") +
  theme_minimal() +
  facet_grid(Gender~Alignment)
```

We would probably visualise the data better with a darker canvas
```{r}
ggplot(heroes) + 
  geom_point(aes(Weight,  Height, colour= Alignment), shape = "square", alpha=0.5) +
  ggtitle ("Heroes: weight vs height by alignment") +
  facet_grid(Gender~Alignment)
```

Now we try facets on the previous boxplot graph
```{r}
ggplot(data = heroes, mapping = aes(Gender, Weight)) +
  geom_boxplot(aes(colour=Gender)) +
  theme_classic()
```

Again let´s split data by alignment
```{r}
ggplot(data = heroes, mapping = aes(Gender, Weight)) +
  geom_boxplot(aes(colour=Gender)) +
  theme_classic() +
  facet_grid(~Alignment)
```

How to export graphs

```{r}
# Store graph in a variable
myplot = ggplot(data = heroes, mapping = aes(Gender, Weight)) +
  geom_boxplot(aes(colour=Gender)) +
  theme_classic() +
  facet_grid(~Alignment) +
  ggtitle ("Weight by Gender&Alignment")

print(myplot) # This way we can print the variable

# Save graph to file (saves last file )
ggsave("myplot.pdf")
ggsave("myplot.png")

# If we want to specify width and height
ggsave("myplot.pdf", width = 10, height = 5)

# If we want to specify a different file name
ggsave("hero_plot.pdf", 
       plot = myplot, width = 10, height = 5)


# Saving graph without ggsave()
pdf("myplot.pdf")
print(myplot)
dev.off()

# From RScript file we can export directly from the outcome window (go to RScript)

```

